Python as a machine learning language: experiences from the scikit-learn project

### 2013-06-04 paper.tex
Use LLNCS documentclass

\documentclass[a4paper,twocolumn]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage[round]{natbib}

% working title, please change
\title{APIs for hyperparameter optimization in scikit-learn}

% Add your name and affiliation here.
% Don't worry too much about the formatting,
% as we'll have to redo that for the LNCS format.
\author{Mathieu Blondel \\
        Graduate School of System Informatics \\
        Kobe University \\
        \small{\texttt{mblondel@ai.cs.kobe-u.ac.jp}}
      \and
        Lars Buitinck \\
        Informatics Institute \\
        University of Amsterdam \\
        \small{\texttt{L.J.Buitinck@uva.nl}}
      \and
        Jake Vanderplas \\
        Astronomy Department \\
        University of Washington \\
        \small{\texttt{vanderplas@astro.washington.edu}}
}

@@

This paper discusses design choices made
in both the programmers' interface (API) and the implementation
of scikit-learn,
with a focus on the problem of hyperparameter optimization.
%It also highlights obstacles faced by users and implementers
%of machine learning algorithms in the Python language.
\end{abstract}

\section{Introduction}

\subsection{Project vision}

@@

The main goal of the scikit-learn project
is to provide high-quality implementations
of ``classic'' machine learning algorithms
with a consistent and clear API.
This paper explores the role that such an API plays
for handling the problem of hyperparameter optimization,
also known as \textit{model selection} (\citealp{elemstatlearn}:219--257).

The rest of this section explains some fundamental design choices
made throughout scikit-learn.
% TODO and the rest of the paper...

@@

%In the remainder of the paper, we use the following notation. Training data are
%denoted by \Xtr ~and test data (i.e., unseen data to which the learning
%algorithm need be able to generalize) by \Xte.  Following the discussion above,
%we represent \Xtr ~and \Xte ~by 2-dimensional NumPy arrays or SciPy sparse
%matrices. Using NumPy/SciPy's notation, \Xtr\texttt{[i, j]} denotes the
%$i^\textrm{th}$ training sample's $j^\textrm{th}$ feature. For supervised learning tasks,
%such as classification or regression, we denote the training and test labels by
%\ytr ~and \yte, respectively. We typically store \ytr ~and \yte ~as
%1-dimensional NumPy arrays (\ytr\texttt{[i]} denotes the $i^\textrm{th}$ training
%sample's label value).

\section{The hyperparameter optimization problem}
\label{hyperparams}

The focus of many machine learning toolkits is the \textit{learning} task,
which, at least in the supervised case,
can be formalized as the problem of finding a function
$\mathsf{model} : \mathcal{X} \to \mathcal{Y}$,
where $\mathcal{X}$ is some set of inputs
and $\mathcal{Y}$ some set of targets
($\mathbb{R}^k$ in regression, $\{1,\dots,K\}$ in classification).
An algorithm that finds such a function is called a \textit{learner}
% TODO: cite Mitchell 1998
or alternatively an \textit{estimator}
(in the scikit-learn documentation, and therefore in this paper).
Using function currying, its function signature is

\[
  L : \mathcal{H} \to \mathcal{X} \to \mathcal{Y}
\]

where $\mathcal{H}$ is the set of all possible \textit{hyperparameter} settings
for the estimator in question.\footnote{
  We disregard the fact that different estimators
  have widely varying sets of hyperparameters for notational simplicity.}
Hyperparameters include, for example,
regularization strength of linear models or SVMs,
choice of loss function for generic linear model learners,
or the number of features to keep in a feature selection method.
They are often treated as user-settable parameters
and not part of the learning problem proper,
either because they would affect the learning problem
in a way that would make it intractable \citep{bergstra2012}
or because their effects cannot properly be measured
in terms of a loss function on a training set.

However, in recent years, algorithms have been published
that systematically explore spaces of possible hyperparameter settings.
The problem of hyperparameter tuning
is crucial to achieving good generalization performance in machine learning.
The present paper does not go into hyperparameter optimization algorithms
in depth, focusing instead on API support for using them.

@@

The bulk of the scikit-learn application programming interface
consists of estimator classes.
All learning algorithms for classification, regression and clustering
are offered as such classes,
but so are feature extraction, feature selection and dimensionality reduction.
The three latter types of estimator are examples of a subtype
called \textit{transformers}:
in this case, both input and output are NumPy arrays representing samples,
but commonly with different numbers of columns.
Transformers that perform feature extraction on e.g.\ free text
are an exception in that their input is not usually a NumPy array.

Construction and application of an estimator object
proceeds in a way that is similar to partial function application,
mirroring the $L : \mathcal{H} \to \mathcal{X} \to \mathcal{Y}$ notation
given for the supervised learning problem in Section~\ref{hyperparams}.
The constructor, \texttt{\_\_init\_\_}\footnote{
  Strictly speaking, \texttt{\_\_init\_\_} is Python's
  class \textit{initialization} method,
  while \texttt{\_\_new\_\_} is the constructor.
  The distinction is not relevant to the present description of scikit-learn.}
takes a set of named (hyper-)parameters
and attaches these to the estimator as public attributes.\footnote{
  The Python language does not actually distinguish between public and private
  attributes and methods,
  but by convention, any identifier prefixed with an underscore (\texttt{\_})
  is considered ``a non-public part of the API\ldots
  an implementation detail and subject to change without notice.''
  \Citep{pythontut}
  }
It never performs actual learning, nor does it even see a training set.

The algorithmic workhorse of most estimators, and one of the few methods
that all estimators have in common, is called \texttt{fit}.
Invoking this method causes the estimator to learn from a training set,
supplied as one or more arguments.
Its task, from an API point of view,
is to determine (via a learning algorithm) model-specific parameters
from a training set and set these as attributes on the estimator object,
which can then be used to transform data or make predictions.
As in other object-oriented machine learning libraries
(e.g.\ Weka's Java API)
no distinction is made in the type hierarchy
between an estimator and the model that results from applying it to data.
When the \texttt{fit} method has been called,
an estimator object serves dual purpose as a model.

As an example to clarify the partial application style,
consider a linear support vector machine (SVM).
% Rewrite example using LinearSVC
If $X$ is a dataset in the format described above and $k$ is some integer,
then the Python code to train a $k$-means clustering model on it
can be written as
% XXX we could use Pygments for syntax highlighting
\begin{verbatim}
km = KMeans(n_clusters=k).fit(X)
\end{verbatim}
This snippet relies on the fact that the \texttt{fit} method
always returns the object it was called on (its ``\texttt{self}''),
making a method chaining style possible.

Importantly, the parameters learned by an estimator
are exposed as \textit{public} attributes on the trained object.
This facilitates model inspection
and makes it possible to train a model using scikit-learn,
export it to a (standardized or custom) external data format
and use the learned parameters in a different piece of code,
perhaps a prediction algorithm implemented in a different language
or one that combines the model parameters with other information
to build a custom model.
Of course, it also forces the developers to think carefully
about what might in a more ``black box''-style toolkit
be considered the internals of objects,
since authors are client code are allowed to rely
on the names and formats of model parameters.

A \texttt{fit} method is present even on objects
that perform seemingly mundane tasks such as vectorizing text documents
for subsequent learning.
Such classes, however, can be said to learn their vocabulary
from the training corpus,
and follow the conventions set out above.
Even those estimators that are entirely stateless and therefore
do not require a \texttt{fit} method to perform useful work,
nevertheless have such a method for consistency and composibility,
and to perform parameter validation.
Examples of this latter kind of objects are the normalizing transformer,
kernel approximation transformers
\citep{rahimi2007random, li2010random, vedaldi2010efficient},
and two feature extraction classes that implement the ``hashing trick''
\citep{weinberger2009}.

\section{Pipelines}

A feature of scikit-learn estimators
that makes effective hyperparameter optimization possible
is the fact that they can be composed into \texttt{Pipeline} objects.
Not only do such objects combine typical machine learning workflows
(extracting features, centering/scaling/normalization,
feature selection, making predictions)
in a single object that is itself an estimator,
and whose methods cause data to flow through the pipeline.
More importantly, though, they can be used for
systematic, automated model selection.
This is best explained by example.

By composing, say, a \texttt{CountVectorizer} object
(which extracts term frequency features from text documents
according to a bag-words model),
a \textsf{tf--idf} transformer, a $\chi^2$ feature selector and a linear SVM,
one obtains a pipeline object with various hyperparameters:
whether to do stopword filtering,
linear vs.\ logarithmic \textsf{tf}, number of features to keep,
SVM regularization ($L_1$ or $L_2$ norm, strength $C$) a.o.
Such a composite \texttt{Pipeline} object
exposes all of these parameters using a special syntax;
each estimators must be given a name,
and if the SVM's name is, e.g., \texttt{linearsvm},
then its $C$ parameter is exposed as \texttt{linearsvm\_\_C}.
The pipeline's \texttt{set\_params} splits the parameter names
on double underscores, uses the first part of the result
to look up the estimator,
and passes the second part to the estimator's own \texttt{set\_params}
as a keyword argument; due to Python's dynamic parameter passing,
the result is the same as if the user had called \texttt{set\_params}
on the SVM themselves.
(The double underscore syntax is not special to pipelines;
it is also employed in ensemble estimators such as random forests.)

The aforementioned \texttt{Pipeline} object
can be passed as an argument to a model selection algorithm.
Available algorithms at the time of writing include random search
\citep{bergstra2012} and ``grid search''.
The latter, which we shall use in this example,
performs an exhaustive search through a grid formed by the Cartesian product
of a set of possible values for each parameter, specified by the user.
In our example, this grid might be given as

@@

At each point in this grid, $k$-fold cross validation is run
to estimate the performance of the estimator according to some measure
(e.g.\ accuracy, $F_\beta$-score or a host of clustering quality metrics)
and the best performing set of hyperparameters is stored
on the \texttt{GridSearchCV} object; again, as a public attribute.

\subsection{Recursive pipelines}

In some cases, it is useful to build recursive pipelines,
and scikit-learn has support for these:
pipelines are themselves estimators,
and can thus be nested inside larger pipelines.

Direct use of this idiom would not have any benefit
over using the individual stages of the sub-pipeline directly,
as both approaches would simply produce a linear pipeline.
However, the recursion support shines in more complex situations
that involve the construction of estimators using \textit{meta-estimators}.
These are objects in the scikit-learn API
that wrap an estimator to give it additional functionality.
An example of meta-estimators are object of the class
\texttt{OneVsRestClassifier},
which can be used to construct a multi-label classifier
\citep{tsoumakas2007multi}
from an ordinary binary classifier object
by constructing one binary classifier per class
(a strategy variously called one-vs.-the-rest,
one-vs.-all or binary relevance):

\begin{verbatim}
ovr_svm = OneVsRestClassifier(LinearSVC())
\end{verbatim}

The \texttt{ovr\_svm} object can be passed to a hyperparameter search
producedure directly to determine parameters such as $C$,
but the syntax to do that has changed.
The object under consideration is no longer a \texttt{LinearSVC}
but a \texttt{OneVsRestClassifier},
which exposes the former object as its public attribute \texttt{estimator}.
Double underscore syntax must be used to address $C$,
i.e.\ \texttt{estimator\_\_C}.

To make this example more interesting,
feature selection using $\chi^2$ tests can be added to the SVM,
such that each of the separate binary classification problems
gets its own set of features.
This can be done by wrapping a pipeline in an OvR object:

\begin{verbatim}
svm_chi2 = Pipeline(["chi2", SelectKBest(chi2)],
                    ["svm", LinearSVC()])
ovr_svm = OneVsRestClassifier(svm_chi2)
\end{verbatim}

Now, the parameter $k$ specifying number of features to be kept
in each $\chi^2$ feature selection step
is exposed by \texttt{ovr\_svm} as \texttt{estimator\_\_chi2\_\_k},
mirroring the nesting of estimators in meta-estimators,
and can be optimized by grid search or random search.

Taking the example one step further to arrive at recursive pipelines,
the \texttt{ovr\_svm} object can be combined
with a \textsf{tf--idf} document vectorizer
to obtain a full document classification pipeline
that takes a stream of text files, paired with class labels, as input:

\begin{verbatim}
doc_clf = Pipeline(["tfidf", TfidfVectorizer()],
                   ["clf", ovr_svm])
\end{verbatim}

The double underscore syntax again serves to address attributes
in this triply nested structure,
so the linear SVM's $C$ parameter
is now called \texttt{clf\_\_estimator\_\_svm\_\_C}.

Following the Python mantra that ``explicit is better than implicit,''
the implementers of scikit-learn have resisted the temptation
to implement shortcuts in the notation,
for example to allow addressing of attributes
of the underlying estimator of the OvR classifier
without the intermediate \texttt{estimator}.
Doing so might require conflict resolution
when an attribute of the meta-estimator object itself
has the same name as an attribute on the wrapped estimator.
The rules for such conflict resolution would need to be
designed, implemented, tested, and, equally important, documented
such as to be understandable for users.
So, the cost of ``syntactic sugar'' was avoided
at the expense of some typing by the user and some loss of elegance.

@@

At this point in time, support for multi-core programming in Python,
at least in the CPython reference implementation targeted by scikit-learn,
is one of the weak spots in the language and its libraries.
The interpreter supports multithreading,
but due to the ``Global Interpreter Lock'' (GIL),
no two threads in a process may at the same time execute Python code.
It follows that multithreading is not a viable solution
even two embarrassingly parallel problems,
such as fitting and evaluating multiple models in a grid search procedure.
The workaround is to use Python's multiprocessing module,
which emulates threads using multiple operating system-level processes,
but this in turn leads to high memory use and overhead
because of the copying of training and test sets
into the various processes' address spaces.\footnote{
  The fact that copying occurs at all is an artifact of NumPy's implementation;
  no attempt at implementing a multiprocessing module
  that shares NumPy arrays using copy-on-write semantics or memory mapping
  has so far been completed.}

>>>>

\documentclass{llncs}

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage[round]{natbib}

% working title, please change
\title{Python as a machine learning language:
       experiences from the scikit-learn project}

\author{Mathieu Blondel\inst{1} \and
        Lars Buitinck\inst{2} \and
        Gilles Louppe\inst{3} \and
        Jake Vanderplas\inst{4}}

\institute{Graduate School of System Informatics, Kobe University \and
           Informatics Institute, University of Amsterdam \and
           Dept. of EE \& CS, University of Liège \and
           Astronomy Department, University of Washington}

@@

This paper discusses design choices made
in both the programmers' interface (API) and the implementation
of scikit-learn.
It also highlights obstacles faced by users and implementers
of machine learning algorithms in the Python language.
\end{abstract}

\section{Project vision}

\subsection{General principles}

@@

The main goal of the scikit-learn project
is to provide high-quality implementations
of ``classic'' machine learning algorithms
with a consistent API.

@@

In the remainder of the paper, we use the following notation. Training data are
denoted by \Xtr ~and test data (i.e., unseen data to which the learning
algorithm needs to be able to generalize) by \Xte.  Following the discussion above,
we represent \Xtr ~and \Xte ~by 2-dimensional NumPy arrays or SciPy sparse
matrices. Using NumPy/SciPy's notation, \Xtr\texttt{[i, j]} denotes the
$i^\textrm{th}$ training sample's $j^\textrm{th}$ feature. For supervised learning tasks,
such as classification or regression, we denote the training and test labels by
\ytr ~and \yte, respectively. We typically store \ytr ~and \yte ~as
1-dimensional NumPy arrays (\ytr\texttt{[i]} denotes the $i^\textrm{th}$ training
sample's label value).

@@

The bulk of the scikit-learn application programming interface
consists of so-called \textit{estimator} objects.
All learning algorithms for classification, regression and clustering
are offered as such objects,
but so are feature extraction, feature selection and dimensionality reduction
algorithms.

Construction and application of an estimator object
proceeds in a way that is somewhat reminiscent of partial function application.
The constructor, \texttt{\_\_init\_\_}\footnote{
  Strictly speaking, \texttt{\_\_init\_\_} is Python's
  class \textit{initialization} method,
  while \texttt{\_\_new\_\_} is the constructor.
  The distinction is not relevant to the present description of scikit-learn.}
takes a set of named (hyper-)parameters
and attaches these to the estimator as public attributes.\footnote{
  The Python language does not actually distinguish between public and private
  attributes and methods,
  but by convention, any identifier prefixed with an underscore (\texttt{\_})
  is considered ``a non-public part of the API\ldots
  an implementation detail and subject to change without notice.''
  \Citep{pythontut}
  }
It never performs actual learning, nor does it even see a training set.

The algorithmic workhorse of most estimators, and one of the few methods
that all estimators have in common, is called \texttt{fit}.
Invoking this method causes the estimator to learn from a training set,
supplied as one or more arguments.
Its task, from an API point of view,
is to determine (via a learning algorithm) model-specific parameters
from a training set and set these as attributes on the estimator object,
which can then be used to transform data or make predictions.
As in other object-oriented machine learning libraries
(e.g.\ Weka's Java API)
no distinction is made in the type hierarchy
between what is variously called an estimator or a \textit{learner}
in the machine learning literature,
and the \textit{model} that results from applying an estimator to data.
When the \texttt{fit} method has been called,
an estimator object serves dual purpose as a model
(and is referred to as such in the documentation).

An example to clarify the partial application style
is the following code snippet.
If $X$ is a dataset in the format described below and $k$ is some integer,
then the Python code to train a $k$-means clustering model on it
can be written as
% XXX we could use Pygments for syntax highlighting
\begin{verbatim}
km = KMeans(n_clusters=k).fit(X)
\end{verbatim}
This snippet relies on the fact that the \texttt{fit} method
always returns the object it was called on (its ``\texttt{self}''),
making a method chaining style possible.

Importantly, the parameters learned by an estimator
are exposed as \textit{public} attributes on the trained object.
This facilitates model inspection
and makes it possible to train a model using scikit-learn,
export it to a (standardized or custom) external data format
and use the learned parameters in a different piece of code,
perhaps a prediction algorithm implemented in a different language
or one that combines the model parameters with other information
to build a custom model.
% XXX Hard to follow phrasing below
Of course, it also forces the developers to think carefully
about what might, in a more ``black box''-style toolkit,
be considered the internals of objects,
since authors of client code are allowed to rely
on the names and formats of model parameters.

A \texttt{fit} method is present even on objects
that perform seemingly mundane tasks such as vectorizing text documents
for subsequent learning.
Such classes, however, can be said to learn their vocabulary
from the training corpus,
and follow the conventions set out above.
Even those estimators that are entirely stateless and therefore
do not require a \texttt{fit} method to perform useful work,
nevertheless have such a method for consistency and composibility,
and to perform parameter validation.
Examples of this latter kind of objects are the normalizing transformer,
kernel approximation transformers
\citep{rahimi2007random, li2010random, vedaldi2010efficient},
% XXX is it relevant that they are exactly two? Propose s/two/the
and two feature extraction classes that implement the ``hashing trick''
\citep{weinberger2009}.

\section{Pipelines and model selection}

A distinguishing feature of scikit-learn estimators
is that they can be composed into \texttt{Pipeline} objects.
Such objects combine typical machine learning workflows
(extracting features, centering/scaling/normalization,
feature selection, making predictions)
in a single object that is itself an estimator,
and whose methods cause data to flow through the pipeline.
More importantly, though, they can be used for
systematic, automated model selection.
This is best explained by example.

By composing, say, a \texttt{CountVectorizer} object
(which extracts term frequency features from text documents
according to a bag-words model),
a \textsf{tf--idf} transformer, a $\chi^2$ feature selector and a linear SVM,
one obtains a pipeline object with various hyperparameters:
whether to do stopword filtering,
linear vs.\ logarithmic \textsf{tf}, number of features to keep,
SVM regularization ($L_1$ or $L_2$ norm, strength $C$) a.o.
Such a composite \texttt{Pipeline} object
exposes all of these parameters using a special syntax;
each estimators must be given a name,
and if the SVM's name is, e.g., \texttt{linearsvm},
then its $C$ parameter is exposed as \texttt{linearsvm\_\_C}.
The pipeline's \texttt{set\_params} splits the parameter names
on double underscores, uses the first part of the result
to look up the estimator,
and passes the second part to the estimator's own \texttt{set\_params}
as a keyword argument; due to Python's dynamic parameter passing,
the result is the same as if the user had called \texttt{set\_params}
on the SVM themselves.
(The double underscore syntax is not special to pipelines;
it is also employed in ensemble estimators such as random forests
% Is the following clear? Suggestions to replace "direct attributes" with"?
It is chosen as a convention to avoid ambiguity, as direct
attributes and methods of scikit-learn objects never contain
double underscores.)

The aforementioned \texttt{Pipeline} object
can be passed as an argument to a model selection algorithm.
Available algorithms at the time of writing include random search
\citep{bergstra2012} and ``grid search''.
The latter, which we shall use in this example,
performs an exhaustive search through a grid formed by the Cartesian product
of a set of possible values for each parameter, specified by the user.
In our example, this grid might be given as:

@@

At each point in this grid, $k$-fold cross validation is run
to estimate the performance of the estimator according to some measure
(e.g.\ accuracy, $F_\beta$-score or a host of clustering quality metrics)
and the best performing set of hyperparameters is stored
on the \texttt{GridSearchCV} object, again, as a public attribute.

@@

At this point in time, support for multi-core programming in Python,
at least in the CPython reference implementation targeted by scikit-learn,
is downright weak.
The interpreter supports multithreading, but due to the ``Global Interpreter Lock'' (GIL),
no two threads in a process may at the same time execute Python code.
It follows that multithreading is not a viable solution
even two embarrassingly parallel problems,
such as fitting and evaluation multiple models in a grid search procedure.
The workaround is to use Python's multiprocessing module,
which emulates threads using multiple operating system-level processes,
but this in turn leads to high memory use and overhead
because of the copying of training and test sets
into the various processes' address spaces.\footnote{
  The fact that copying occurs at all is an artifact of NumPy's implementation;
  no attempt at implementing a multiprocessing module
  that shares NumPy arrays using copy-on-write semantics or memory mapping
  has so far been completed.}
